---
title: "Spatial Autocorrelation in Ecological Data"
author: "Vojta Barták"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required packages

```{r}
library(tidyverse)
library(sf)
library(stars)
library(tmap)
library(spdep)
library(gstat)
#library(geoR)
```

## Loading and visualizing data

```{r}
krnap <- read.table("data/krnap_richness.csv", sep=",", header=T) %>% 
  st_as_sf(coords=c("X", "Y"), crs=5514, remove=F)
dtm <- read_stars("data/dtm_CR_1000.tif")
```

```{r}
tmap_mode("view")
qtm(krnap)
```

```{r}
tm_shape(krnap) +
  tm_dots("richness")
```

```{r}
tmap_mode("plot")
tm_shape(dtm) +
  tm_raster(style="cont", palette="Blues") +
  tm_shape(krnap, is.master = T) +
  tm_dots("richness", size=1) +
  tm_layout(legend.outside = T)
```

## Measuring and testing global spatial autocorrelation

```{r}
neibs <- dnearneigh(st_coordinates(krnap), d1=1000, d2=2000)
plot(neibs, coords=st_coordinates(krnap))
```

```{r}
weights <- nb2listw(neibs, style="B")
summary(weights)
summary(unlist(weights$weights))
```

graphical comparison of the richness and the weighted average of it's neighbors:

```{r}
mp <- moran.plot(krnap$richness, weights)
krnap$infl <- mp$is_inf
tm_shape(krnap) +
  tm_dots("infl", size = 1)
mp <- moran.plot(krnap$richness, nb2listw(neibs, style="W"))
krnap$infl <- mp$is_inf
tm_shape(krnap) +
  tm_dots("infl", size = 1)
```



```{r}
moran.test(krnap$richness, weights)
```

try randomization = F
try different weighting style?
try greary.test

```{r}
lm.morantest(lm(richness~1, krnap), weights)
```

there exists a saddlepoint approximation, and an exact combinatorial test:

```{r}
lm.morantest.exact(lm(richness~1, krnap), weights)
```
and if the sample size is too large, a Monte Carlo test as well:

```{r}
moran.mc(krnap$richness, weights, nsim=999)
```

distance-based weights:

```{r}
dst <- nbdists(neibs, st_coordinates(krnap))
idw <- lapply(dst, function(x) 1/x)
dst.weights <- nb2listw(neibs, glist = idw, style="B")
summary(unlist(dst.weights$weights))
```

```{r}
moran.test(krnap$richness, dst.weights)
```

## Local spatial autocorrelation

```{r}
loc.mor <- localmoran(krnap$richness, nb2listw(neibs, style = "W"))
head(loc.mor)
```
```{r}
loc.mor <- localmoran.exact(lm(richness~1, krnap), nb = neibs)
head(as.data.frame(loc.mor))
```

```{r}
krnap$local_I <- sapply(loc.mor, function(x) x$estimate)
krnap$local_p <- sapply(loc.mor, function(x) x$p.value)
tm_shape(krnap) +
  tm_dots(c("local_I","local_p"), size=1) +
  tm_layout(legend.position = c("LEFT","BOTTOM"))
```

## Correlogram, Variogram, Covariogram

```{r}
sp.cor <- sp.correlogram(neibs, krnap$richness, order=10)
sp.cor
```

```{r}
plot(sp.cor)
```

```{r}
sp.cor <- sp.correlogram(neibs, krnap$richness, order=10, method = "I")
sp.cor
```

```{r}
plot(sp.cor)
```
```{r}
library(pgirmess)
sp.cor2 <- correlog(st_coordinates(krnap), krnap$richness, method = "Moran")
sp.cor2
```

```{r}
plot(sp.cor2)
```

try different nbclass..

```{r}
variogram(richness~1, data = krnap, cutoff = 40000, cloud=T) %>% plot
```

```{r}
vario <- variogram(richness~1, cutoff = 40000, data = krnap)
plot(vario)
```

```{r}
covario <- variogram(richness~1, data = krnap, cutoff=40000, covariogram=T)
tail(covario)
```

```{r}
plot(covario)
```
```{r}
plot(covario$dist, covario$gamma/max(covario$gamma))
```

```{r}
data.frame(distance = covario$dist, correlation = covario$gamma/max(covario$gamma)) %>% 
  arrange(desc(correlation)) %>% 
  "["(2:nrow(.),) %>% 
  ggplot(aes(x=distance, y=correlation)) +
  geom_line() +
  geom_point()
```

## Residual autocorrelation

```{r}
krnap$elev <- st_extract(dtm, krnap) %>% pull(1)
plot(richness~elev, krnap)
abline(lm(richness~elev, krnap))
```
```{r}
ggplot(krnap, aes(x=elev, y=richness)) +
  geom_point() +
  geom_smooth(method="lm", se=T)
```

```{r}
krnap$residual[!is.na(krnap$elev)] <- lm(richness~elev, krnap)$residuals
tm_shape(dtm) +
  tm_raster(style="cont", palette="Blues") +
  tm_shape(krnap, is.master = T) +
  tm_bubbles("residual")
```

```{r}
krnap.subset <- na.omit(krnap)
neibs.res <- dnearneigh(st_coordinates(krnap.subset), d1=1000, d2=2000)
weights.res <- nb2listw(sp.nb.res, style="B")
```


```{r}
moran.plot(krnap.subset$residual, weights.res)
```


```{r}
model <- lm(richness~elev, krnap.subset)
lm.morantest(model, weights.res)
```

```{r}
lm.morantest.exact(model, weights.res)
```


```{r}
loc.mor.res <- localmoran.exact(model=model, nb=neibs.res)
krnap.subset$local_I <- sapply(lm.exact.res, function(x) x$estimate)
krnap.subset$local_p <- sapply(lm.exact.res, function(x) x$p.value)
tm_shape(krnap.subset) +
  tm_dots(c("local_I", "local_p"), size=1) +
  tm_layout(legend.position = c("LEFT","BOTTOM"))
```

```{r}
correlog(st_coordinates(krnap.subset), krnap.subset$residual, method = "Moran") %>% plot
```

```{r}
variogram(richness~elev, data=krnap.subset, cutoff = 40000) %>% plot
```

