---
title: "Spatial Autocorrelation in Ecological Data"
author: "Vojta Barták"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Required packages

```{r}
library(tidyverse)
library(sf)
library(stars)
library(tmap)
library(spdep)
library(gstat)
#library(geoR)
```

## Loading and visualizing data

```{r}
krnap <- read.table("data/krnap_richness.csv", sep=",", header=T) %>% 
  st_as_sf(coords=c("X", "Y"), crs=5514, remove=F)
dtm <- read_stars("data/dtm_CR_1000.tif")
```

```{r}
tmap_mode("view")
qtm(krnap)
```

```{r}
tm_shape(krnap) +
  tm_dots("richness")
```

```{r}
tmap_mode("plot")
tm_shape(dtm) +
  tm_raster(style="cont", palette="Blues") +
  tm_shape(krnap, is.master = T) +
  tm_dots("richness", size=1) +
  tm_layout(legend.outside = T)
```

## Measuring and testing global spatial autocorrelation

```{r}
sp.nb <- dnearneigh(st_coordinates(krnap), d1=1000, d2=2000)
plot(sp.nb, coords=st_coordinates(krnap))
```

```{r}
weights <- nb2listw(sp.nb, style="B")
summary(weights)
summary(unlist(weights$weights))
```

graphical comparison of the richness and the weighted average of it's neighbors:

```{r}
mp <- moran.plot(krnap$richness, weights)
krnap$infl <- mp$is_inf
tm_shape(krnap) +
  tm_dots("infl", size = 1)
mp <- moran.plot(krnap$richness, nb2listw(sp.nb, style="W"))
krnap$infl <- mp$is_inf
tm_shape(krnap) +
  tm_dots("infl", size = 1)
```



```{r}
moran.test(krnap$richness, weights)
```

try randomization = F
try different weighting style?
try greary.test

```{r}
lm.morantest(lm(richness~1, krnap), weights)
```

there exists a saddlepoint approximation, and an exact combinatorial test:

```{r}
lm.morantest.exact(lm(richness~1, krnap), weights)
```
and if the sample size is too large, a Monte Carlo test as well:

```{r}
moran.mc(krnap$richness, weights, nsim=999)
```

distance-based weights:

```{r}
dst <- nbdists(sp.nb, st_coordinates(krnap))
idw <- lapply(dst, function(x) 1/x)
dst.weights <- nb2listw(sp.nb, glist = idw, style="B")
summary(unlist(dst.weights$weights))
```

```{r}
moran.test(krnap$richness, dst.weights)
```

## Local spatial autocorrelation

```{r}
loc.mor <- localmoran(krnap$richness, nb2listw(sp.nb, style = "W"))
head(loc.mor)
```
```{r}
loc.mor <- localmoran.exact(lm(richness~1, krnap), nb = sp.nb)
head(as.data.frame(loc.mor))
```

```{r}
krnap$local_I <- sapply(loc.mor, function(x) x$estimate)
krnap$local_p <- sapply(loc.mor, function(x) x$p.value)
tm_shape(krnap) +
  tm_dots(c("local_I","local_p"), size=1) +
  tm_layout(legend.position = c("LEFT","BOTTOM"))
```

## Correlogram, Variogram, Covariogram

```{r}
sp.cor <- sp.correlogram(sp.nb, krnap$richness, order=10)
sp.cor
```

```{r}
plot(sp.cor)
```

```{r}
sp.cor <- sp.correlogram(sp.nb, krnap$richness, order=10, method = "I")
sp.cor
```

```{r}
plot(sp.cor)
```
```{r}
library(pgirmess)
sp.cor2 <- correlog(st_coordinates(krnap), krnap$richness, method = "Moran")
sp.cor2
```

```{r}
plot(sp.cor2)
```

try different nbclass..

```{r}
variogram(richness~1, data = krnap, cutoff = 40000, cloud=T) %>% plot
```

```{r}
vario <- variogram(richness~1, cutoff = 40000, data = krnap)
plot(vario)
```

